% ====================================
% ACKNOWLEDGEMENT PAGE
% ====================================

\thispagestyle{plain}

\begin{center}
{\fontsize{18}{19}\selectfont \textbf{Acknowledgment}}
\end{center}

\vspace{1.2cm}

\begin{spacing}{1.3}

\noindent
I extend my heartfelt gratitude to my supervisor, \textbf{Dr.\ Ravinder Singh}, for his unwavering support, steadfast guidance, and invaluable mentorship during the dissertation phase. I am thankful for his constructive suggestions and encouragement throughout the dissertation work. His belief in my capabilities has been a powerful source of motivation throughout this process.

\vspace{0.4cm}

\noindent
I am also grateful to express my special thanks to \textbf{Dr.\ Vinay Kumar}, Head of the Department of STATISTICS, Central University of Haryana, for providing me an opportunity to complete my work. His profound expertise, visionary leadership, and continuous encouragement have significantly shaped my academic journey and bestowed confidence in my capabilities.

\vspace{0.4cm}

\noindent
I would like to express my deepest gratitude to all the faculty members of the \\Department of STATISTICS for their good wishes, inspiration, and continuous support throughout my dissertation.

\vspace{0.4cm}

\noindent
I am grateful to research scholars, my peers, and colleagues for their collaboration, insightful discussions, and unwavering support throughout this project. Their contribution was crucial in overcoming obstacles and achieving milestones.

\vspace{0.4cm}

\noindent
I am appreciative of my family's support, tolerance, and faith in my abilities. Their affection and encouragement have been the driving force behind my efforts in academia.

\end{spacing}

\vspace{2cm}

\begin{flushright}
\textbf{Name: } Vishal Verma\\[0.2cm]
\textbf{Roll No:} 240959 \\[0.2cm]
\textbf{Session:} 2025--26 \\[0.2cm]
\textbf{Reg. No:}  24033240959\\[0.2cm]
Department of  \\
Central University of Haryana, Mahendergarh
\end{flushright}

\cleardoublepage
